---
title: "Step-specific selection"
output:
  html_document:
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, tidy = 'styler')
library(data.table)
library(ggplot2)
tutorial_dir = system.file("step_selection_tutorial", package = 'cancereffectsizeR')
```

# Introduction

`ces_variant()` estimates a single scaled selection coefficient for each variant, assuming
selection is constant across every sample included in the analysis. Sometimes, though, your
samples represent an ordered sequence of tumor progression stages — normal tissue and primary
tumor, say, or primary tumor and metastasis, or more than two stages — and you'd like to know
whether (and how) selection on a variant changes from one stage to the next.

`ces_variant_step()` and its supporting functions extend cancereffectsizeR to this setting: instead
of one selection intensity per variant, you get one per stage, estimated jointly from a single
likelihood model. This vignette walks through the workflow: assigning samples to stages,
calculating each gene's per-stage mutation rate proportions, running the model, testing whether a
step-specific model actually fits better than a plain `ces_variant()` run, and plotting the
results.

This is a substantial modeling exercise, so it's worth restating the core assumption plainly: your
samples must represent a real, ordered progression (each later stage having passed through every
earlier one), and each variant you analyze must belong to exactly one gene. If your samples are
better described as independent subgroups (not a progression), a stratified `ces_variant()`
analysis (running separately per subgroup) is more appropriate.

# Assigning samples to stages

The first ingredient is a `sample_index` table associating every sample with an ordered stage.
`assign_stage_index()` builds it from a sample-level metadata column (added via `load_maf()` or
`load_sample_data()`) and a description of stage order:

```{r}
cesa = assign_stage_index(cesa, stage_col = "tissue_type",
                          stage_order = c("Normal", "Tumor"))
```

`stage_order` just needs to be in earliest-to-latest order. For more than two stages, or when
multiple raw values should be grouped into the same stage, use a named list instead:

```{r}
sample_index = assign_stage_index(cesa, stage_col = "tissue_type",
                                  stage_order = list(Normal = "Normal",
                                                     Primary = "Primary_tumor",
                                                     Metastasis = c("Lymph_node_met", "Distant_met")))
```

If you already have a table associating samples with stages, you can build `sample_index` by hand
instead: it just needs `Unique_Patient_Identifier`, `group_index` (1 = earliest stage), and
`group_name` columns.

# Calculating stage-specific mutation rate proportions

The model also needs, for each gene, what proportion of its total mutation rate is estimated to
accumulate during each stage. This comes from running `gene_mutation_rates()` once per stage,
each time restricted to the samples that have reached at least that stage (so a later stage's
mutation rate reflects a cumulative total, having passed through every earlier stage first):

```{r}
cesa = gene_mutation_rates(cesa, covariates = "your_tissue_type",
                           samples = cesa$samples[tissue_type == "Normal"])
cesa = gene_mutation_rates(cesa, covariates = "your_tissue_type",
                           samples = cesa$samples[tissue_type == "Tumor"])
```

Each call adds a new `rate_grp_<N>` column to `get_gene_rates(cesa)`. `stage_mutation_proportions()`
converts these cumulative, per-stage rates into per-gene proportions:

```{r}
stage_mut_prop = stage_mutation_proportions(cesa, rate_cols = c("rate_grp_1", "rate_grp_2"),
                                            stage_names = c("Normal", "Tumor"))
```

```{r, echo=FALSE, eval=TRUE}
stage_mut_prop = readRDS(system.file("step_selection_tutorial/kras_stage_mut_prop.rds", package = "cancereffectsizeR"))
```
```{r, eval=TRUE}
stage_mut_prop
```

Cumulative rates should never decrease from one stage to the next (a later stage covers at least
as much mutational time as an earlier one), but dN/dS-based rate estimates are noisy, and a small
number of genes can come out non-monotonic just from that noise. By default
(`on_invalid = "floor"`), affected genes get a warning and have their negative/undefined stage
contribution floored to a tiny share (`floor_prop`, default 1e-6) of the gene's total rate, with
proportions rescaled to still sum to 1. Set `on_invalid = "NA"` to instead leave those genes' rows
as `NA` (so you can decide how to handle them, or drop them from `variants` before running
`ces_variant_step()`), or `on_invalid = "error"` to stop and inspect them yourself.

# Running the model

With `sample_index` and `stage_mut_prop` in hand, `ces_variant_step()` works much like
`ces_variant()`:

```{r}
kras_variants = select_variants(cesa, genes = "KRAS", min_freq = 2)
cesa = ces_variant_step(cesa, variants = kras_variants, stage_mut_prop = stage_mut_prop,
                        sample_index = sample_index, run_name = "step_effects",
                        return_fit = TRUE, conf = 0.95)
```

(You can pass `stage_col`/`stage_order` directly to `ces_variant_step()` instead of a pre-built
`sample_index`, exactly as with `assign_stage_index()`.)

The output table has one selection intensity column per stage (`si_<stage name>`), and, when
`conf` is set, matching confidence interval columns:

```{r, echo=FALSE, eval=TRUE}
step_effects = readRDS(system.file("step_selection_tutorial/kras_step_effects.rds", package = "cancereffectsizeR"))
```
```{r, eval=TRUE}
step_effects
```

Every variant (or compound variant) passed to `ces_variant_step()` must map to exactly one gene
present in `stage_mut_prop` — the mutation-rate proportions are gene-level, so an intergenic
variant or a compound spanning multiple genes has no single row to use.

# Is the step-specific model actually better?

A step-specific model always fits at least as well as a simple `ces_variant()` model, because it
has more free parameters (one per stage, instead of one overall). To see whether the improvement
is more than you'd expect by chance, run the simple model too and compare with a likelihood ratio
test:

```{r}
cesa = ces_variant(cesa, variants = kras_variants, run_name = "simple_effects", return_fit = TRUE)
lrt = step_selection_LRT(cesa, step_run_name = "step_effects", simple_run_name = "simple_effects")
```

```{r, echo=FALSE, eval=TRUE}
lrt = readRDS(system.file("step_selection_tutorial/kras_step_lrt.rds", package = "cancereffectsizeR"))
```
```{r, eval=TRUE}
lrt
```

# Plotting results

`plot_effects_step()` visualizes step-specific effects, with one panel per variant (or gene, or
any other grouping column), stage on the x-axis, and points colored by stage. It works for any
number of stages. Passing the `step_selection_LRT()` output via `lrt` adds a significance
annotation to each panel.

```{r}
plot_effects_step(step_effects, group_by = "variant_name", lrt = lrt,
                  stage_order = c("Normal", "Tumor"))
```

```{r, echo=FALSE, eval=TRUE, fig.width=8, fig.height=5}
readRDS(system.file("step_selection_tutorial/kras_step_plot.rds", package = "cancereffectsizeR"))
```

Note that a stage's confidence interval can come out `NA` on one side when that stage's fitted
selection intensity sits right at the optimizer's boundary (as with KRAS G12D's `Normal`-stage
estimate above, which hit the 0.001 lower bound) — that's an accurate reflection of a genuinely
undefined bound there, not a bug, and `plot_effects_step()` simply omits that side of the error
bar.
